A performance comparison of Dask and Apache Spark for data-intensive neuroimaging pipelines

### 2019-07-29 latex/paper.tex
Minor edits in introduction

\author{Mathieu Dugr\'e, Val\'erie Hayot-Sasson, Tristan Glatard\\
Department of Computer Science and Software Engineering\\
Concordia University, Montr\'eal, Qu\'ebec, Canada
}

@@

\section{Introduction}
% Context
The recent rise in data sharing and improved data collection strategies
have brought neuroimaging to the Big Data era ~\cite{ALFAROALMAGRO:18,
UKBioBank:18}. Existing neuroimaging workflow engines, such as
Nipype~\cite{Nipype:11}, are well suited for processing the standard
compute-intensive neuroimaging pipelines, but lack incorporated Big Data
strategies (i.e. in-memory computing, data locality and lazy-evaluation) to
improve performance of the increasingly prevalent data-intensive pipelines.
As was noted in ~\cite{hayot2019performance}, in-memory computing, coupled
with data locality, can bring significant performance improvements to data
intensive neuroimaging pipelines. We extend this work by studying the
differences between two Big Data frameworks, Dask~\cite{Dask:15} and Apache
Spark~\cite{Spark:16}, for their suitability in the processing of neuroimaging
pipelines. Our goal is to test whether Spark or Dask has a clear
performance advantage to process Big neuroimaging Data.

Spark and Dask both offer in-memory computing, data locality, and lazy evaluation;
which is common for Big Data frameworks. Both their schedulers operate dynamically,
which is good when task runtimes are not known ahead of time~\cite{Dask:15}. They
also provide rich, high-level programming APIs, and support a variety of
infrastructure schedulers, such as Mesos~\cite{hindman2011mesos}, YARN~\cite{vavilapalli2013apache}, or HPC clusters (Dask only). Over
these similarities, the frameworks have differences.

First and foremost, Spark is written in Scala while Dask is in Python.
Given the popularity of Python in scientific communities, this arguably
gives an edge to Dask due to data serialization costs from Python to Scala.
On the other hand, Python's Global Interpreter Lock (GIL) might reduce
parallelism in some cases. This difference in programming languages also
has qualitative implications. As part of the \href{http://scipy.org}{SciPy}
ecosystem, Dask provides almost transparent parallelization of applications
manipulating Numpy arrays or Pandas data frames. On the other hand, Spark's
Java, R and Python APIs allow to easily parallelize analyses that combine
these languages, with reduced performance loss.

Spark and Dask were both included in the evaluation reported
in~\cite{Mehta:17} that used a neuroimaging and an astronomy application
processing approximately \SI{100}{\giga\byte} of data each. In this work, Dask was
reported to have a slight performance advantage over Spark. Overall, Dask's
end-to-end time (makespan) was measured to be up to 14\% faster than Spark,
due to ``more efficient pipelining'' %(Fig 10) 
and serialization time to
Python. % (Fig. 12-a)
Dask, however, was reported to have a larger startup
time than Spark. % (Fig. 12-c)
The analysis remains at a quite high level
though, leaving most of the observed performance difference unexplained. In
comparison, our study will provide a detailed analysis of performance
differences and similarities between Spark and Dask.

The next section details the design of our experiments. We consider two
data-intensive neuroimaging applications: high-resolution imaging,
represented by the Big Brain data~\cite{Amunts:13}, and large functional
MRI studies, represented by data from the consortium for reliability and
reproducibility (CoRR~\cite{zuo2014open}). We test application pipelines
involving different patterns (map-only, map-reduce), and different types of
implementations (plain Python, command-line,
containerized). We evaluate performance on a dedicated cluster
representative of the ones used in today's data-intensive neuroimaging
studies, using the main Dask and Spark APIs. The following sections present
our results, discussion and conclusions.

@@

\begin{figure}[!b]
    \centering
    \begin{subfigure}[b]{\columnwidth}
        \href{https://zenodo.org/record/3354434/files/spark-inc-baseline.html?download=1}{
            \includegraphics[clip,width=\columnwidth,
            height=0.15\textheight]{images/spark_inc_gantt.png}}
        \caption{Spark}\label{fig:inc_spark_gantt}
    \end{subfigure}
    \\
    \begin{subfigure}[b]{\columnwidth}
        \href{https://zenodo.org/record/3354434/files/dask-bag-inc-baseline.html?download=1}{
        \includegraphics[clip,width=\columnwidth,
        height=0.15\textheight]{images/bag_inc_gantt.png}}
        \caption{Dask Bag}\label{fig:inc_dask_bag_gantt}
    \end{subfigure}
    \\
    \begin{subfigure}[b]{\columnwidth}
        \href{https://zenodo.org/record/3354434/files/dask-delayed-inc-baseline.html?download=1}{
        \includegraphics[clip,width=\columnwidth,
        height=0.15\textheight]{images/delayed_inc_gantt.png}}
        \caption{Dask Delayed}\label{fig:inc_dask_delayed_gantt}
    \end{subfigure}
    \\
    \begin{subfigure}[b]{\columnwidth}
        \href{https://zenodo.org/record/3354434/files/dask-futures-inc-baseline.html?download=1}{
        \includegraphics[clip,width=\columnwidth,
        height=0.15\textheight]{images/futures_inc_gantt.png}}
        \caption{Dask Futures}\label{fig:inc_dask_futures_gantt}
    \end{subfigure}
    \caption{Incrementation: 125 blocks, 10 iterations, \SI{4}{\second} delay, 8 workers,
8 CPUs/worker}\label{fig:inc_gantt}
\end{figure}

@@

\begin{figure}[!t]
    \centering
    \begin{subfigure}[b]{\columnwidth}
        \href{https://zenodo.org/record/3354434/files/spark-histo-baseline.html?download=1}{
        \includegraphics[clip,width=\columnwidth,
        height=0.15\textheight]{images/spark_histo_gantt.png}}
        \caption{Spark}\label{fig:histo_spark_gantt}
    \end{subfigure}
    \\
    \begin{subfigure}[b]{\columnwidth}
        \href{https://zenodo.org/record/3354434/files/dask-bag-histo-baseline.html?download=1}{
        \includegraphics[clip,width=\columnwidth,
        height=0.15\textheight]{images/bag_histo_gantt.png}}
        \caption{Dask Bag}\label{fig:histo_dask_bag_gantt}
    \end{subfigure}
    \\
    \begin{subfigure}[b]{\columnwidth}
        \href{https://zenodo.org/record/3354434/files/dask-delayed-histo-baseline.html?download=1}{
        \includegraphics[clip,width=\columnwidth,
        height=0.15\textheight]{images/delayed_histo_gantt.png}}
        \caption{Dask Delayed}\label{fig:histo_dask_delayed_gantt}
    \end{subfigure}
    \caption{Histogram: 125 blocks, 8 workers, 8
    CPUs/worker}\label{fig:histo_gantt}
\end{figure}

@@

\begin{figure}[!tb]
    \centering
    \begin{subfigure}[b]{\columnwidth}
        \href{https://zenodo.org/record/3354434/files/spark-histo_np-baseline.html?download=1}{
        \includegraphics[clip,width=\columnwidth,
        height=0.15\textheight]{images/spark_histo_np_gantt.png}}
        \caption{Spark}\label{fig:histo_np_spark_gantt}
    \end{subfigure}
    \\
    \begin{subfigure}[b]{\columnwidth}
        \href{https://zenodo.org/record/3354434/files/dask-bag-histo_np-baseline.html?download=1}{
        \includegraphics[clip,width=\columnwidth,
        height=0.15\textheight]{images/bag_histo_np_gantt.png}}
        \caption{Dask Bag}\label{fig:histo_np_dask_bag_gantt}
    \end{subfigure}
    \\
    \begin{subfigure}[b]{\columnwidth}
        \href{https://zenodo.org/record/3354434/files/dask-delayed-histo_np-baseline.html?download=1}{
        \includegraphics[clip,width=\columnwidth,
        height=0.15\textheight]{images/delayed_histo_np_gantt.png}}
        \caption{Dask Delayed}\label{fig:histo_np_dask_delayed_gantt}
    \end{subfigure}
    \caption{Histogram (NumPy): 125 blocks, 8 workers, 8
    CPUs/worker}\label{fig:histo_np_gantt}
\end{figure}

@@

\begin{figure}[!htb]
    \centering
    \begin{subfigure}[b]{\columnwidth}
        \href{https://zenodo.org/record/3354434/files/spark-bids-baseline.html?download=1}{
        \includegraphics[clip,width=\columnwidth,
        height=0.15\textheight]{images/spark_bids_gantt.png}}
        \caption{Spark}\label{fig:bids_spark_gantt}
    \end{subfigure}
    \\
    \begin{subfigure}[b]{\columnwidth}
        \href{https://zenodo.org/record/3354434/files/dask-bag-bids-baseline.html?download=1}{
        \includegraphics[clip,width=\columnwidth,
        height=0.15\textheight]{images/bag_bids_gantt.png}}
        \caption{Dask Bag}\label{fig:bids_dask_bag_gantt}
    \end{subfigure}
    \\
    \begin{subfigure}[b]{\columnwidth}
        \href{https://zenodo.org/record/3354434/files/dask-delayed-bids-baseline.html?download=1}{
        \includegraphics[clip,width=\columnwidth,
        height=0.15\textheight]{images/delayed_bids_gantt.png}}
        \caption{Dask Delayed}\label{fig:bids_dask_delayed_gantt}
    \end{subfigure}
    \\
    \begin{subfigure}[b]{\columnwidth}
        \href{https://zenodo.org/record/3354434/files/dask-futures-bids-baseline.html?download=1}{
        \includegraphics[clip,width=\columnwidth,
        height=0.15\textheight]{images/futures_bids_gantt.png}}
        \caption{Dask Futures}\label{fig:bids_dask_futures_gantt}
    \end{subfigure}
    \caption{BIDS App example: 8 workers}\label{fig:bids_gantt}
\end{figure}

@@

\subsection{Effect of data-intensive applications}
In our experiments, Incrementation, Histogram (NumPy) and Histogram for Spark have
low compute time while BIDS App example and Histogram for Dask have significantly
longer compute time. When adding additional workers, we observe that the highly
data-intensive applications scale poorly while compute-intensive applications scale
almost linearly This is due to the limitation from the NFS, as mentioned above.

@@

\subsection{Effect of tasks' size}
We measured the effect of various block sizes for constant BigBrain image for
Incrementation, Histogram, and Histogram (NumPy). We observe that larger block size
results in more IO compensated by lower overhead time and vice-versa.

@@

Varying the block size or the size of the partitions is equivalent. No difference is
observed when varying the task size while using an NFS. However, if one uses a
parallel shared file system, we suppose it would be preferable to use a larger block
size as the IO time should be reduced substantially. It is worth noting that the Dask
developers suggest, in
\href{https://docs.dask.org/en/latest/best-practices.html}{Dask's best practices}, to
use a partition size such that workers have 2-3 times as many chunks available to
work on. More experiments would have to be performed to gain insight on an optimal
number of partitions to use for a given set of data-intensive applications.

@@

\section{Conclusion}
We presented a comparison of two Big Data engines: Spark and Dask. We studied the
engines on three data-intensive neuroimaging applications. Our results show no
substantial difference between the engines. Moreover, the majority of the execution
time was spent on data transfer and overhead. We also observed a possible correlation
between IO and overhead time: an higher overhead is compensated by a lower IO time and
a low overhead is compensated by an high IO time. This would be due to concurrent IO
tasks saturating the bandwidth. We suggest that current schedulers policies would benefit
from knowing the task types, either IO or compute, it is processing concurrently.

>>>>

\author{Mathieu Dugr\'e, Val\'erie Hayot-Sasson, Tristan Glatard\\
Department of Computer Science and Software Engineering\\
Concordia University, Montr\'eal, Qu\'ebec, Canada 
\vspace*{0.8cm}
}

@@

\section{Introduction}
% Context
The recent rise in data sharing and improved data collection strategies
have brought neuroimaging to the Big Data era ~\cite{ALFAROALMAGRO:18,
van2014human}. Existing neuroimaging workflow engines, such as
Nipype~\cite{Nipype:11}, are well suited for processing the standard
compute-intensive neuroimaging pipelines, but lack incorporated Big Data
strategies (i.e. in-memory computing, data locality and lazy-evaluation) to
improve performance of the increasingly prevalent data-intensive pipelines.
As was noted in ~\cite{hayot2019performance}, in-memory computing, coupled
with data locality, can bring significant performance improvements to data
intensive neuroimaging pipelines. We extend this work by studying the
differences between two Big Data frameworks, Dask~\cite{Dask:15} and Apache
Spark~\cite{Spark:16}, for their suitability in the processing of neuroimaging
pipelines. Our goal is to test whether Spark or Dask has a clear
performance advantage to process Big neuroimaging Data.

%Similarities
Spark and Dask both offer in-memory computing, data locality, and lazy evaluation,
which is common for Big Data frameworks. Both their schedulers operate dynamically,
which is good when task runtimes are not known ahead of time~\cite{Dask:15}. They
also provide rich, high-level programming APIs, and support a variety of
infrastructure schedulers, such as Mesos~\cite{hindman2011mesos}, YARN~\cite{vavilapalli2013apache}, or HPC clusters (Dask only). Over
these similarities, the frameworks have differences.

% Differences
First and foremost, Spark is written in Scala while Dask is in Python.
Given the popularity of Python in scientific communities, this arguably
gives an edge to Dask due to data serialization costs from Python to Scala.
On the other hand, Python's Global Interpreter Lock (GIL) might reduce
parallelism in some cases. This difference in programming languages also
has qualitative implications. As part of the \href{http://scipy.org}{SciPy}
ecosystem, Dask provides almost transparent parallelization of applications
manipulating Numpy arrays or Pandas data frames. On the other hand, Spark's
Java, R and Python APIs allow to easily parallelize analyses that combine
these languages, with reduced performance loss. Our study focuses on
performance, although we recognize that other factors will also impact the
choice of a framework in practice.

% Related work
Spark and Dask were both included in the evaluation reported
in~\cite{Mehta:17}, where a neuroimaging application
processed approximately \SI{100}{\giga\byte} of data. In this work, Dask was
reported to have a slight performance advantage over Spark. Overall, Dask's
end-to-end time (makespan) was measured to be up to 14\% faster than Spark,
due to ``more efficient pipelining'' %(Fig 10) 
and serialization time to
Python. % (Fig. 12-a)
Dask, however, was reported to have a larger startup
time than Spark. % (Fig. 12-c)
The analysis remained at a quite high level
though, leaving most of the observed performance difference unexplained. In
comparison, our study will provide a detailed analysis of performance
differences and similarities between Spark and Dask for neuroimaging data processing.

The next section details the design of our benchmarking experiments. We consider two
data-intensive neuroimaging applications: high-resolution imaging,
represented by the Big Brain data~\cite{Amunts:13}, and large functional
MRI studies, represented by data from the consortium for reliability and
reproducibility (CoRR~\cite{zuo2014open}). We test application pipelines
involving different patterns (map-only, map-reduce), and different types of
implementations (plain Python, command-line,
containerized). We evaluate performance on a dedicated cluster
representative of the ones used in today's data-intensive neuroimaging
studies, using the main Dask and Spark APIs. The other sections present
our results, discussion and conclusions.

@@

\begin{figure}[!b]
    \centering
    \begin{subfigure}[b]{\columnwidth}
        \includegraphics[clip,width=\columnwidth,
        height=0.15\textheight]{images/spark_inc_gantt.png}
        \caption{Spark}\label{fig:inc_spark_gantt}
    \end{subfigure}
    \\
    \begin{subfigure}[b]{\columnwidth}
        \includegraphics[clip,width=\columnwidth,
        height=0.15\textheight]{images/bag_inc_gantt.png}%
        \caption{Dask Bag}\label{fig:inc_dask_bag_gantt}
    \end{subfigure}
    \\
    \begin{subfigure}[b]{\columnwidth}
        \includegraphics[clip,width=\columnwidth,
        height=0.15\textheight]{images/delayed_inc_gantt.png}%
        \caption{Dask Delayed}\label{fig:inc_dask_delayed_gantt}
    \end{subfigure}
    \\
    \begin{subfigure}[b]{\columnwidth}
        \includegraphics[clip,width=\columnwidth,
        height=0.15\textheight]{images/futures_inc_gantt.png}%
        \caption{Dask Futures}\label{fig:inc_dask_futures_gantt}
    \end{subfigure}
    \caption{Incrementation: 125 blocks, 10 iterations, \SI{4}{\second} delay, 8 workers,
8 CPUs/worker}\label{fig:inc_gantt}
\end{figure}

@@

\begin{figure}[!t]
    \centering
    \begin{subfigure}[b]{\columnwidth}
        \includegraphics[clip,width=\columnwidth,
        height=0.15\textheight]{images/spark_histo_gantt.png}
        \caption{Spark}\label{fig:histo_spark_gantt}
    \end{subfigure}
    \\
    \begin{subfigure}[b]{\columnwidth}
        \includegraphics[clip,width=\columnwidth,
        height=0.15\textheight]{images/bag_histo_gantt.png}%
        \caption{Dask Bag}\label{fig:histo_dask_bag_gantt}
    \end{subfigure}
    \\
    \begin{subfigure}[b]{\columnwidth}
        \includegraphics[clip,width=\columnwidth,
        height=0.15\textheight]{images/delayed_histo_gantt.png}%
        \caption{Dask Delayed}\label{fig:histo_dask_delayed_gantt}
    \end{subfigure}
    \caption{Histogram: 125 blocks, 8 workers, 8
    CPUs/worker}\label{fig:histo_gantt}
\end{figure}

@@

\begin{figure}[!tb]
    \centering
    \begin{subfigure}[b]{\columnwidth}
        \includegraphics[clip,width=\columnwidth,
        height=0.15\textheight]{images/spark_histo_np_gantt.png}
        \caption{Spark}\label{fig:histo_np_spark_gantt}
    \end{subfigure}
    \\
    \begin{subfigure}[b]{\columnwidth}
        \includegraphics[clip,width=\columnwidth,
        height=0.15\textheight]{images/bag_histo_np_gantt.png}%
        \caption{Dask Bag}\label{fig:histo_np_dask_bag_gantt}
    \end{subfigure}
    \\
    \begin{subfigure}[b]{\columnwidth}
        \includegraphics[clip,width=\columnwidth,
        height=0.15\textheight]{images/delayed_histo_np_gantt.png}%
        \caption{Dask Delayed}\label{fig:histo_np_dask_delayed_gantt}
    \end{subfigure}
    \caption{Histogram (NumPy): 125 blocks, 8 workers, 8
    CPUs/worker}\label{fig:histo_np_gantt}
\end{figure}

@@

\begin{figure}[!htb]
    \centering
    \begin{subfigure}[b]{\columnwidth}
        \includegraphics[clip,width=\columnwidth,
        height=0.15\textheight]{images/spark_bids_gantt.png}
        \caption{Spark}\label{fig:bids_spark_gantt}
    \end{subfigure}
    \\
    \begin{subfigure}[b]{\columnwidth}
        \includegraphics[clip,width=\columnwidth,
        height=0.15\textheight]{images/bag_bids_gantt.png}%
        \caption{Dask Bag}\label{fig:bids_dask_bag_gantt}
    \end{subfigure}
    \\
    \begin{subfigure}[b]{\columnwidth}
        \includegraphics[clip,width=\columnwidth,
        height=0.15\textheight]{images/delayed_bids_gantt.png}%
        \caption{Dask Delayed}\label{fig:bids_dask_delayed_gantt}
    \end{subfigure}
    \\
    \begin{subfigure}[b]{\columnwidth}
        \includegraphics[clip,width=\columnwidth,
        height=0.15\textheight]{images/futures_bids_gantt.png}%
        \caption{Dask Futures}\label{fig:bids_dask_futures_gantt}
    \end{subfigure}
    \caption{BIDS App example: 8 workers}\label{fig:bids_gantt}
\end{figure}

@@

\subsection{Effect of data-intensive applications}
In our experiments, Incrementation, Histogram (NumPy) and Histogram for Spark have
low compute time while BIDS App example and Histogram for Dask have significantly
longer compute time. When adding additional workers, we observe that data-intensive
applications scale poorly while compute-intensive applications scale almost linearly
This is due to the limitation from the NFS, as mentioned above.

@@

\subsection{Effect of tasks' size}
We measured the effect of various block sizes for constant BigBrain image for
Incrementation, Histogram, and Histogram (NumPy). We observe that larger block size
results in more IO compensated by lower overhead time. On the other hand, smaller
block size results in higher overhead time compensated by reduced IO time.

@@

Varying the block size is equivalent to varying the size of the partitions. We
observed no difference when varying the task size while using an NFS. However, if
one uses a parallel shared file system, we suppose it would be preferable to use a
larger block size as the IO time should be reduced substantially. It is worth noting
that the Dask developers suggest, in
\href{https://docs.dask.org/en/latest/best-practices.html}{Dask's best practices}, to
use a partition size such that workers have 2-3 times as many chunks available to
work on. More experiments would have to be performed to gain insight on an optimal
number of partitions to use for a given set of data-intensive applications.

@@
